# Random variables {#random-variables}

```{r setup4, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      prompt = FALSE,
                      tidy = TRUE,
                      collapse = TRUE)
library("tidyverse")
```

Economics is a mostly *quantitative* field: its outcomes can be usually
described using numbers: price, quantity, interest rates, unemployment rates,
GDP, etc. Statistics is also a quantitative field: every statistic is a number
calculated from data.  When a random outcome is described by a number, we call
that number a ***random variable***.  We can use probability theory to describe
and model random variables.

This chapter will introduce the basic terminology and mathematical tools for
working with simple random variables.

::: {.goals data-latex=""}
**Chapter goals**

In this chapter, we will learn how to:

1.  Define a random variable in terms of a random outcome.
2.  Determine the support and range of a random variable.
3.  Calculate and interpret the PDF of a discrete random variable.
4.  Calculate and interpret the CDF of a discrete random variable.
5.  Calculate interval probabilities from the CDF.
6.  Calculate the expected value of a discrete random variable from its PDF.
7.  Calculate a quantile from the CDF.
8.  Calculate the variance of a discrete random variable from its PDF.
9.  Calculate the variance from expected values.
9.  Calculate the standard deviation from the variance.
10. Calculate the expected value for a linear function of a random variable.
11. Calculate the variance and standard deviation for a linear function of a
    random variable.
12. Standardize a random variable.
13. Use standard discrete probability distributions:
    - Bernoulli
    - binomial
    - discrete uniform.
:::

To prepare for this chapter, please review the chapter on
[probability and random events](#probability) and the section on 
[sequences and summations](#sequences-and-summations) in the math appendix.

## Defining a random variable {#introduction-to-random-variables}

A ***random variable*** is a number whose value depends on a random outcome. The
idea here is that we are going to use a random variable to describe some (but
not necessarily every) aspect of the outcome.

::: example
**Random variables in roulette**

Here are a few random variables we could define in a roulette game:

- The number the ball lands on: 
  $$b = b(\omega) = \omega$$
- An indicator for whether a bet on red wins:
  $$r = r(\omega) = I(\omega \in RedWins)=\begin{cases}1 & \textrm{if } \omega \in RedWins \\ 0 & \textrm{if } \omega \notin RedWins \\ \end{cases}$$
- The player's net profits from \$1 bet on red:
      $$ y_{Red} = y_{Red}(\omega) = \begin{cases}  1 & \textrm{ if } \omega \in RedWins \\ -1 & \textrm{ if } \omega \in RedWins^c \end{cases} $$
      That is, a player who bets \$1 on red wins \$1 if the ball lands on red 
      and loses \$1 if the ball lands anywhere else.
- The player's net profits from \$1 bet on 14:
      $$ y_{14} = y_{14}(\omega) = \begin{cases}  35 & \textrm{ if } \omega = 14 \\ -1 & \textrm{ if } \omega \neq 14 \end{cases} $$
      That is, a player who bets \$1 on 14 wins \$35 if the ball lands on 14 
      and loses \$1 if the ball lands anywhere else.

All of these random variables are defined in terms of the underlying outcome 
$\omega$.
:::

A random variable is always a function of the original outcome, but for
convenience, we usually leave its dependence on the original outcome implicit,
and write it as if it were an ordinary variable.

### Implied distribution {#probability-distributions}

Since every random variable is a number, we can define its sample space as the
set of real numbers $\mathbb{R}$.

Each random variable has its own probability distribution over this sample space
and this probability distribution can be derived from the probability
distribution of the underlying outcome.  That is, let
$\omega \in \Omega$ be some random outcome, and let $x = x(\omega)$
be some random variable that depends on that outcome.  Then the probability
that $x$ is in some set $A$ is:
$$\Pr(x \in A) = \Pr(\{\omega \in \Omega: x(\omega) \in A\})$$
Again, this definition looks complicated but is easier to follow with a few
simple examples.

::: example
**Probability distributions for roulette**

Assume we have a fair roulette game.  Since each of our random variables is a
function of the outcome $\omega$, we can derive their probability distributions
from the probability distribution of $\omega$:

   - The probability distribution for $b$ is:
      $$\Pr(b = 0) = \Pr(\omega = 0) = 1/37 \approx 0.027$$
      $$\Pr(b = 1) = \Pr(\omega = 1) = 1/37 \approx 0.027$$
      $$\vdots$$
      $$\Pr(b = 36) = \Pr(\omega = 36) = 1/37 \approx 0.027$$
      $$\Pr(b = 37) = \Pr(\omega \notin \{0,1,\ldots,36\}) = 0$$
   - The probability distribution for $r$ is:
      $$\Pr(r = 1) = \Pr(\omega \in RedWins) = 18/37 \approx 0.486$$
      $$\Pr(r = 0) = \Pr(\omega \notin RedWins) = 19/37 \approx 0.514$$
      $$\Pr(r \notin \{0,1\}) = 0$$
   - The probability distribution for $y_{14}$ is:
      $$\Pr(y_{14} = 35) = \Pr(\omega = 14) = 1/37 \approx 0.027$$
      $$\Pr(y_{14} = -1) = \Pr(\omega \neq 14) = 36/37 \approx 0.973$$
      $$\Pr(y_{14} \notin \{-1,35\}) = 0$$

Notice that these random variables are related to each other since they all
depend on the same underlying outcome $\omega$. Section \@ref(multiple-random-variables)
will explain how we can describe and analyze those relationships.
:::

### The support {#the-support}

The ***support*** of a random variable $x$ is the smallest[^401] set
$S_x \subset \mathbb{R}$ such that $\Pr(x \in S_x) = 1$.

[^401]: Technically, it is the smallest *closed* set, but let's ignore that.

In plain language, the support is the set of all values in the sample space that
have some chance of actually happening.

::: example
**The support in roulette**

The sample space of $b$ is $\mathbb{R}$ and the support of $b$ is
$S_{b} = \{0,1,2,\ldots,36\}$.

The sample space of $r$ is $\mathbb{R}$ and the support of
$r$ is $S_{r} = \{0,1\}$.

The sample space of $y_{14}$ is $\mathbb{R}$ and the support of
$y_{14}$ is $S_{14} = \{-1,35\}$.
:::

The random variables we will consider in this chapter  have  ***discrete***
support. That is, the support is a set of isolated points each of which has
a strictly positive probability. In most examples the support will also have
a ***finite*** number of elements.  All finite sets are also discrete, but
it is also possible for a discrete set to have an infinite number of elements.
For example, the set of positive integers $\{1,2,3,\ldots\}$ is both discrete
and infinite.

Some random variables have a support that is continuous rather than discrete.
Chapter \@ref(more-on-random-variables) will cover continuous random variables. 

## The PDF and CDF {#the-pdf-and-cdf}

### The PDF {#the-pdf-of-a-discrete-random-variable}

We can describe the probability distribution of a random variable with a
function called its ***probability density function (PDF)***.

The PDF of a discrete random variable $x$ is defined as:
  $$f_x(a) = \Pr(x = a)$$
where $a$ is any number (the value of $a$ does *not* need to be in the support).
By convention, we typically use a lower-case $f$ to represent a PDF, and we
use the subscript when needed to clarify which specific random variable we are
talking about.

The probability distribution and PDF of a random variable are closely related
but distinct. The probability distribution is a function that gives the
probability $\Pr(x \in A)$ for any event (set) $A$, while the PDF is a function
that gives the gives the probability $\Pr(x = a)$ for any number $a$.

You can derive the PDF from the probability distribution.

::: example
**The PDF in roulette**

The random variables $b$, $r$ and $y_{14}$ are all discrete, and each has its
own PDF:

\begin{align}
f_b(a) &= \Pr(b = a) = \begin{cases}
          1/37 & \textrm{if } a \in \{0,1,\ldots,36\} \\
          0 & \textrm{otherwise} \\
        \end{cases} \\
f_{r}(a) &= \Pr(r = a) = \begin{cases}
            19/37 & \textrm{if } a = 0 \\
            18/37 & \textrm{if } a = 1 \\
            0 & \textrm{otherwise}  \\
          \end{cases} \\
f_{14}(a) &= \Pr(y_{14} = a) = \begin{cases}
            36/37 & \textrm{if } a = -1 \\
            1/37 & \textrm{if } a = 35 \\
            0 & \textrm{otherwise} \\
          \end{cases} \\
\end{align}
Figure \@ref(fig:RoulettePDF) below shows these three PDFs.

```{r RoulettePDF, fig.cap = "*PDFs for the roulette example*"}
RoulettePDF <- tibble(a = seq(from = -2, to= 36),
                      fb = c(0, 0, rep(1/37, times = 37)),
                      fr = c(0, 0, 19/37, 18/37, rep(0, times=35)),
                      fred = c(0, 19/37, 0, 18/37, rep(0, times=35)),
                      f14 = c(0, 36/37,rep(0, times=35), 1/37, 0))
ggplot(data = RoulettePDF, mapping = aes(x = a)) +
  geom_point(aes(y=fb), col = "blue") +
  geom_point(aes(y=fr), col = "red") +
  geom_point(aes(y=f14), col = "orange") +
  xlab("a") +
  ylab("f(a)") +
  ylim(0,1) +
  geom_text(x = 4, 
            y = 4/37, 
            label = "f_b(a)", 
            col = "blue") +
  geom_text(x = 4, 
            y = 18/37, 
            label = "f_r(a)", 
            col = "red") +
  geom_text(x = 2, 
            y = 36/37, 
            label = "f_14(a)", 
            col = "orange") +
  labs(title = "Probability density function (PDF)", 
       subtitle = "Roulette", 
       caption = "", 
       tag = "")
```

Note that the points overlap, so you may not be able to see each value for a
given PDF.
:::

You can also derive the probability distribution from the PDF. For any discrete
random variable $x$ and any event $A \subset \mathbb{R}$ we can calculate 
$\Pr(x \in A)$ by simply adding up the corresponding PDF of $x$:
  \begin{align}
    \Pr(x \in A) &= \sum_{s \in A} \Pr(x = s) \\
                 &= \sum_{s \in S_x} f_x(s)I(s \in A)
  \end{align}
If you are unfamiliar with the notation here, please refer to the sections on
[summations](#summations) and [the indicator function](#the-indicator-function)
in the Math Review Appendix. The formula is easy to use once you understand the
notation.

::: example
**Some event probabilities in roulette**

Since the outcome in roulette is discrete, we can calculate any event
probability by adding up the probabilities of the event's component outcomes.

The probability of the event $b \leq 1$ can be calculated:
  \begin{align}
    \Pr(b \leq 1) &= \sum_{s=0}^{36}f_b(s)I(s \leq 1) \\
      &= \underbrace{f_b(0)}_{1/37} \underbrace{I(0 \leq 1)}_{1} +
          \underbrace{f_b(1)}_{1/37} \underbrace{I(1 \leq 1)}_{1} +
          \underbrace{f_b(2)}_{1/37} \underbrace{I(2 \leq 1)}_{0} +
          \cdots +
          \underbrace{f_b(36)}_{1/37} \underbrace{I(36 \leq 1)}_{0} \\
      &= 2/37
  \end{align}
  
The probability of the event $b \in Even$ can be calculated:
  \begin{align}
    \Pr(b \in Even) &= \sum_{s=0}^{36}f_b(s)I(s \in Even) \\
      &= \underbrace{f_b(0)}_{1/37} \underbrace{I(0 \in Even)}_{0} +
          \underbrace{f_b(1)}_{1/37} \underbrace{I(1 \in Even)}_{0} +
          \underbrace{f_b(2)}_{1/37} \underbrace{I(2 \in Even)}_{1} +
          \cdots +
          \underbrace{f_b(36)}_{1/37} \underbrace{I(36 \in Even)}_{1} \\
        &= 18/37
  \end{align}
Remember that zero is not counted as an even number in roulette, so it is not
in the event $Even$.
:::

The PDF of a discrete random variable has several general properties:

1. It is always between zero and one:
    $$0 \leq f_x(a) \leq 1$$
   since it is a probability.
2. It sums up to one over the support:
    $$\sum_{a \in S_x} f_x(a) = \Pr(x \in S_x) = 1$$
   since the support has probability one by definition.
3. It is strictly positive for all values in the support:
    $$a \in S_x \implies f_x(a) > 0$$
   since the support is the *smallest* set that has probability one.

You can confirm that examples above all satisfy these properties.

### The CDF {#the-cdf}

Another way to describe the probability distribution of a random variable is
with a function called its ***cumulative distribution function (CDF)***. 
The CDF is a little less intuitive than the PDF, but it has the advantage that
it always has the same definition whether the random variable is discrete,
continuous, or even some combination of the two.

The CDF of the random variable $x$ is the function 
$F_x:\mathbb{R} \rightarrow [0,1]$ defined by:
  $$F_x(a) = Pr(x \leq a)$$
where $a$ is any number.  By convention, we typically use an upper-case $F$ to
indicate a CDF, and we use the subscript to indicate what random variable we are
talking about.

We can construct the CDF of a discrete random variable by just adding up the
PDF.  For any value for $a$, we can calculate:
  \begin{align}
    F_x(a) &= \Pr(x \leq a) \\
      &= \sum_{s \in S_x} f_x(s)I(s \leq a)
  \end{align}
This formula might look intimidating, but it is actually quite easy to
implement.

::: example
**Deriving the CDF of $y_{14}$**

We will derive the CDF $F_{14}(\cdot)$ of the random variable $y_{14}$
using the formula above and following this step-by-step procedure:

1. *Write down the formula*, substituting in the appropriate variable names:
   \begin{align}
     F_{14}(a) &= \Pr(y_{14} \leq a) \\
      &= \sum_{s \in S_{14}} f_{14}(s)I(s \leq a)
   \end{align}
2. *Find the support and substitute it into the formula*. In this case, we
   earlier found that the support of $y_{14}$ is $S_{14} = \{-1, 35\}$ so:
   \begin{align}
     F_{14}(a) &= \sum_{s \in \{-1, 35\} } f_{14}(s)I(s \leq a)
   \end{align}
3. *Expand out the summation*:
   \begin{align}
     F_{14}(a) &= f_{14}(-1)I(-1 \leq a) + f_{14}(35)I(35 \leq a)
   \end{align}
   If you don't know how to do this, review the material on summations
   in the Math Appendix.
4. *Find the PDF and substitute it into the formula*. In this case, we earlier
   found that $f_{14}(-1) = 36/37 \approx 0.973$ and
   $f_{14}(35) = 1/37 \approx 0.027$, so:
   \begin{align}
     F_{14}(a) &= 36/37 * I(-1 \leq a) + 1/37 * I(35 \leq a) \\
      &\approx 0.973 * I(-1 \leq a) + 0.027 * I(1 \leq a) \\
   \end{align}
5. *Try out a few values for $a$*. You can try as many values as you like until
   you get a feel for what this function looks like:
   \begin{align}
     F_{14}(-100) &\approx 0.973 * \underbrace{I (-1 \leq -100)}_{=0} + 0.027 * \underbrace{I(35 \leq -100)}_{=0} \\
      &= 0 \\
     F_{14}(-2) &\approx 0.973 * \underbrace{I (-1 \leq -2)}_{=0} + 0.027 * \underbrace{I(35 \leq -2)}_{=0} \\
      &= 0 \\
     F_{14}(-1) &\approx 0.973 * \underbrace{I (-1 \leq -1)}_{=1} + 0.027 * \underbrace{I(35 \leq -1)}_{=0} \\
      &\approx 0.973 \\
     F_{14}(0) &\approx 0.973 * \underbrace{I (-1 \leq 0)}_{=1} + 0.027 * \underbrace{I(35 \leq 0)}_{=0} \\
      &\approx 0.973 \\
     F_{14}(35) &\approx 0.973 * \underbrace{I (-1 \leq 35)}_{=1} + 0.027 * \underbrace{I(35 \leq 35)}_{=1} \\
      &= 1 \\
     F_{14}(100) &\approx 0.973 * \underbrace{I (-1 \leq 100)}_{=1} + 0.027 * \underbrace{I(35 \leq 100)}_{=1} \\
      &= 1 
   \end{align}
6. *Summarize your results in a clear and simple formula*. Case notation is a
   good way to do this:
   \begin{align}
     F_{14}(a) &\approx \begin{cases}
                  0 & a < -1 \\
                  0.973 & -1 \leq a < 35 \\
                  1 & a \geq 35 \\
                  \end{cases}
   \end{align}

Figure \@ref(fig:RouletteCDF1) below shows this CDF. As you can see, it starts
off at zero, jumps up at the values $-1$ and $35$ and then stays at one after
that.

```{r RouletteCDF1, fig.cap = "*A CDF for the roulette example*"}
RouletteCDF <- RoulettePDF %>%
    mutate (Fb = cumsum(fb)) %>%
    mutate (Fred = cumsum(fred))%>%
    mutate (F14 = cumsum(f14)) 
ggplot(data = RouletteCDF, mapping = aes(x = a)) +
  geom_step(aes(y=F14), col = "orange") +
  xlab("a") +
  ylab("F(a)") +
  geom_text(x = 8, 
            y = 32/37, 
            label = "F_14(a)",
            col = "orange") +
  labs(title = "Cumulative distribution function (CDF)", 
       subtitle = "Roulette - net player profits for a bet on 14", 
       caption = "", 
       tag = "")
```

  
:::

I have given a specific formula and step-by-step instructions here, but once 
you have done a few of these calculations, you will probably be able to handle
most cases without explicitly following each step. If you are stuck, come back
to these instructions.

::: example
**More CDFs for roulette**

We can follow the same procedure to derive CDFs for the other random variables:

  - The CDF of $b$ is:
     $$F_b(a) = \begin{cases}
                   0 & a < 0 \\
                   1/37 & 0 \leq a < 1 \\
                   2/37 & 1 \leq a < 2 \\
                   \vdots & \vdots \\
                   36/37 & 35 \leq a < 36 \\
                   1 & a \geq 36 \\
                \end{cases}$$
  - The CDF of $y_{14}$ is:
    $$F_{14}(a) = \begin{cases}
                0 & a < -1 \\
                36/37 & -1 \leq a < 35 \\
                1 & a \geq 35 \\
                \end{cases}$$
:::

If you plot a CDF you will notice it follows certain patterns.

::: example
**CDF properties**

Figure \@ref(fig:RouletteCDF) below graphs the CDFs from the previous example:  

```{r RouletteCDF, fig.cap = "*CDFs for the roulette example*"}
RouletteCDF <- RoulettePDF %>%
    mutate (Fb = cumsum(fb)) %>%
    mutate (Fr = cumsum(fr))%>%
    mutate (Fred = cumsum(fred))%>%
    mutate (F14 = cumsum(f14)) 
ggplot(data = RouletteCDF, mapping = aes(x = a)) +
  geom_step(aes(y=Fb), col = "blue") +
  geom_step(aes(y=Fr), col = "red") +
  xlab("a") +
  ylab("F(a)") +
  geom_text(x = 7, 
            y = 4/37, 
            label = "F_b(a)",
            col = "blue") +
  geom_text(x = 3, 
            y = 18/37, 
            label = "F_r(a)",
            col = "red") +
  labs(title = "Cumulative distribution function (CDF)", 
       subtitle = "Roulette - number ball lands on (b), red wins indicator (r) ", 
       caption = "", 
       tag = "")
```

Notice that they show a few common properties:

  - The CDF runs from zero to one, and never leaves that range.
  - The CDF never goes down, it only goes up or stays the same.
  - The CDF has a distinctive "stair-step" shape, jumping up at each point in
    the support, and flat between those points,  
    
In fact, the CDF of *any* discrete random variable has these properties.
:::

All CDFs have the following properties:

1. The CDF is a *probability,*  That is:
     $$0 \leq F_x(a) \leq 1$$
    for any number $a$.
2. The CDF is *non-decreasing*.  That is:
     $$F_x(a) \leq F_x(b)$$
   for any two numbers $a$ and $b$ such that $a \leq b$.
3. The CDF  *runs from zero to one*.  That is, it is zero or close to zero for
   low values of $a$, and one or close to one for high values of $a$. We can use
   limits to give precise meaning to the broad terms "close", "low", and "high":
   \begin{align}
     \lim_{a \rightarrow -\infty} F_x(a) = \Pr(x \leq -\infty) &= 0 \\
     \lim_{a \rightarrow \infty} F_x(a) = \Pr(x \leq \infty) &= 1
   \end{align}
   You can review the section on [limits](#limits) in the math appendix if you
   do not follow the notation.
   
In addition, all discrete random variables have the stair-step shape.

In addition to constructing the CDF from the PDF, we can also go the other way,
and construct the PDF of a discrete random variable from its CDF. Each little
jump in the CDF is a point in the support, and the size of the jump is exactly
equal to the PDF.

::: {.fyi data-latex=""}
In more formal mathematics, the formula for deriving the PDF of a discrete
random variable from its CDF would be written:
  $$f_x(a) = F_x(a) - \lim_{\epsilon \rightarrow 0} F_x(a-|\epsilon|)$$
but we can just think of it as the size of the jump.
:::

### Interval probabilities {#interval-probabilities}

We can use the CDF to calculate the probability that $x$ lies in any interval.
That is, let $L$ and $H$ be any two numbers such that $L < H$. Then:
  \begin{align}
    F(H) - F(L) &= \Pr(x \leq H) - \Pr(x \leq L) \\
      &= \Pr((x \leq L) \cup (L < x \leq H)) - \Pr(x \leq L) \\
      &= \Pr(x \leq L) + \Pr(L < x \leq H) - \Pr(x \leq L) \\
      &= \Pr(L < x \leq H) 
  \end{align}
Notice that we have to be a little careful here to distinguish between the
strict inequality $<$ and the weak inequality $\leq$, because it is always
possible for $x$ to be exactly equal to $L$ or $H$. We can use the PDF as
needed to adjust for those possibilities:
\begin{align}
  \Pr(L < x \leq H) &= F(H) - F(L) \\
  \Pr(L \leq x \leq H) &= \Pr(x = L) + \Pr(L < x \leq H) \\
    &= f(a) + F(b) - F(a) \\
  \Pr(L < x < H) &= \Pr(L < x \leq H) - \Pr(x = H) \\
    &= F(H) - F(L) - f(H) \\
  \Pr(L \leq x < H) &= \Pr(x = L) + \Pr(L < x \leq H) - \Pr(x = H) \\
    &= f(L) + F(H) - F(L) - f(H) 
\end{align}

::: example
**Calculating interval probabilities**

Consider the CDF for $b$ derived above. Then:
  \begin{align}
    \Pr(b \leq 36) &= F_b(36) \\
      &= 1 \\
    \Pr(1 < b \leq 36) &= F_b(36) - F_b(1) \\
      &= 1 - 2/37 \\
      &= 35/37
  \end{align}
Note that the placement of the $<$ and $\leq$ are important here.

What if we want $\Pr(1 \leq b \leq 36)$ instead?  We can split that event into
two disjoint events $(b = 1)$ and $(1 < b \leq 36)$ and apply the axioms of
probability:
  \begin{align}
    \Pr(1 \leq b \leq 36) &= \Pr( (b = 1) \cup (1 < b \leq 36) )  \\
      &= \Pr(b = 1) + \Pr(1 < b \leq 36)  \\
      &= f_b(1) + F_b(36) - F_b(1)  \\
      &= 1/37 + 1 - 2/37  \\
      &= 36/37
  \end{align}
We can use similar methods to determine $\Pr(1 < b < 36)$ or
$\Pr(1 \leq b < 36)$.
:::

### Range and mode {#range}

The ***range*** of a random variable is the interval from its lowest possible
value to its highest possible value.  It can be caclulated from the support:
\begin{align}
  range(x) &= [\min(S_x), \max(S_x)]
\end{align}
or from the PDF or CDF.

:::example
**The range in roulette**

The support of $y_{Red}$ is $\{-1,1\}$ so its range is $[-1,1]$.

The support of $y_{14}$ is $\{-1,35\}$ so its range is $[-1,35]$.

The support of $b$ is $\{0,1,2,\ldots,36\}$ so its range is $[0,36]$.
:::

The ***mode*** of a discrete random variable is (loosely speaking) its most
likely value. That is, it is the number $a$ that maximizes $\Pr(x=a)$. The
mathematical language for that would be:
\begin{align}
  mode(x) &= \textrm{arg }\max_{a \in S_x} f_x(a)
\end{align}
A random variable can have more than one mode.

:::example
**The mode in roulette**

The mode of $r$ is $0$, since $f_{r}(0) = 0.514 > 0.486 = f_{r}(1)$.

The mode of $y_{Red}$ is $-1$, since $f_{Red}(-1) = 0.514 > 0.486 = f_{Red}(1)$.

The mode of $y_{14}$ is $-1$, since $f_{14}(-1) = 0.973 > 0.027 = f_{14}(35)$.

The random variable $b$ has 37 modes: $(0,1,2,\ldots,36)$ since each of those
values has the same probability.
:::

## The expected value {#the-expected-value}

The ***expected value*** of a random variable $x$ is written $E(x)$. When $x$
is discrete, it is defined as:

$$E(x) = \sum_{a \in S_x} a\Pr(x=a) = \sum_{a \in S_x} af_x(a)$$
The expected value is also called the ***mean***, the ***population mean***
or the ***expectation*** of the random variable.

The formula for the expected value may look difficult if you are not used to the
notation, but it is actually quite simple to calculate: just multiply each
possible value by its probability, and then add it all up.

::: example
**Calculating the expected value of $y_{14}$**

In previous examples, we found the support and PDF of $y_{14}$.  We can
find its expected value by following these steps:

1. *Rewrite the formula with the correct variable names*:
   \begin{align}
      E(y_{14}) &= \sum_{a \in S_{14}} a f_{14}(a)
   \end{align}
2. *Find the support and substitute*:
   \begin{align}
      E(y_{14}) &= \sum_{a \in \{-1,35\}} a f_{14}(a)
   \end{align}
3. *Expand the summation*:
   \begin{align}
      E(y_{14}) &= -1 * f_{14}(-1) + 35 * f_{14}(35)
   \end{align}
4. *Find the PDF values and substitute*:
   \begin{align}
      E(y_{14}) &= -1 * 36/37  + 35 * 1/37 \\
        &= -1/37 \\
        &\approx -0.027
   \end{align}

:::

The expected value of a random variable is an important concept, and has several
closely related interpretations:

- As a ***weighted average*** of its possible values,  That is, it is like an
  average but instead of each value receiving equal weight it receives a weight
  equal to the probability that we will observe that value.
- As a measure of ***central tendency***, i.e., a typical or representative
  value for the random variable.  Other measures of central tendency include
  the [median](#median) and the [mode](#range).
- As a ***prediction*** for the value of the random variable.  That is, we want
  to predict/guess its value, and we want our guess to be as close as possible
  to the actual (not-yet-known) value.  The median and mode are also used for
  prediction; the best prediction method depends on what you mean by "as close
  as possible":
  - The mode is is the most likely to be *exactly* correct (prediction error
    of zero).
  - The median produces the lowest *absolute* value of the prediction error.
  - The expected value produces the lowest *squared* value of the prediction
     error.

::: example
**Interpreting the expected value of $y_{14}$**

As described earlier, we can think of the expected value as describing a
typical or predicted value for the random variable.  In this case, our
result can be interpreted as a prediction of the house advantage: the player
can expect to lose approximately 2.7 cents per dollar bet on 14, and the house
can expect to gain approximately 2.7 cents per dollar bet on 14.
:::

We can follow the same procedures to find and interpret the expected value of
any discrete random variable.

::: example
**More expected values in roulette**

The support of $b$ is $\{0,1,2\ldots,36\}$ and its PDF is the $f_b(\cdot)$
function we calculated earlier. So its expected value is:
  \begin{align}
    E(b) &= \sum_{a \in S_b} a f_b(a) \\
      &= \sum_{a \in \{0,1,2\ldots,36\}} a f_b(a) \\
      &= 0*\underbrace{f_b(0)}_{1/37} + 1*\underbrace{f_b(1)}_{1/37} + \cdots 36*\underbrace{f_b(36)}_{1/37} \\
      &= \frac{1 + 2 + \cdots + 36}{37} \\
      &= 18
  \end{align}

The support of $r$ is $\{0,1\}$ and its PDF is the $f_r(\cdot)$ function we
calculated earlier. So its expected value is:
  \begin{align}
    E(r) &= \sum_{a \in S_r} a f_r(a) \\
      &= \sum_{a \in \{0,1\}} a f_r(a) \\
      &= 0*\underbrace{f_r(0)}_{19/37} + 1*\underbrace{f_r(1)}_{18/37} \\
      &= 18/37 \\
      &\approx 0.486
  \end{align}
:::

## Quantiles and their relatives {#the-properties-of-a-random-variable}

You have probably heard of the median, and you may have heard of percentiles.
These are special cases of a group of numbers called the ***quantiles***
of a distribution.

### Quantiles and percentiles {#quantiles-and-percentiles}

Let $q$ be any number *strictly* between zero and one.  Then the $q$
***quantile*** of a random variable $x$ is defined as:
  \begin{align}
    F_x^{-1}(q) &= \min\{a \in S_X: \Pr(x \leq a) \geq q\} \\
      &= \min\{a \in S_x: F_x(a) \geq q\}
  \end{align}
where $F_x(\cdot)$ is the CDF of $x$.  The quantile function $F_x^{-1}(\cdot)$
is also called the ***inverse CDF***.

The $q$ quantile of a distribution is also called the $100q$ ***percentile***;
for example the 0.25 quantile of $x$ is also called the 25th percentile of $x$.

::: example
**Quantiles in roulette**

To find quantiles of the random variable $r$, we start by writing down its CDF:
  $$F_{r}(a) = \begin{cases}0 & a < 0 \\
                              19/37 \approx 0.514 & 0 \leq a < 1 \\
                              1 & a \geq 1 \\ \end{cases}$$
Then we plot the CDF.

```{r RouletteQuantiles, fig.cap = "*Deriving quantiles from the CDF*"}
ggplot(data = RouletteCDF, mapping = aes(x = a)) +
  geom_step(aes(y=Fr), col = "red") +
  xlab("a") +
  ylab("F(a)") +
  geom_text(x = 20, 
            y = 0.9, 
            label = "F_r(a)", 
            col = "red") +
  geom_text(mapping=aes(x = 4, y = 0.25),
            label = "0.25 quantile = 0",
            col = "blue") +
  geom_text(x = 5, 
            y = 0.75, 
            label = "0.75 quantile = 1",
            col = "blue") +
  geom_segment(x=-5,xend=0,y=0.25,yend=0.25,col="blue",linetype = "dashed") +
  geom_segment(x=-5,xend=1,y=0.75,yend=0.75,col="blue",linetype = "dashed") +
  labs(title = "Cumulative distribution function (CDF)", 
       subtitle = "Roulette - indicator for red wins (r)", 
       caption = "", 
       tag = "")
```
To find any quantile $q$, we can apply the definition, or just need to find the
value on the graph where $F_{r}(\cdot)$ crosses $q$.

For example, the 0.25 quantile (25th percentile) is defined as:
  \begin{align}
    F_{r}^{-1}(0.25) &= \min\{a \in S_x: F_{r}(a) \geq 0.25\} \\
      &= \min \{0, 1\} \\
      &= 0
  \end{align}
or we can draw the blue dashed line marked "0.25 quantile" and see that it hits
the red line at $a = 0$.

By the same method, we can find that the 0.75 quantile (75th percentile) by
seeing that the red line crosses the blue dashed line marked "0.75 quantile" at
$a = 1$, or we can apply the definition:
  \begin{align}
    F_{r}^{-1}(0.75) &= \min\{a \in S_x: F_{r}(a) \geq 0.75\} \\
      &= \min \{1\} \\
      &= 1
  \end{align}
Either method will work.
:::

The formula for the quantile function may look intimidating, but it can be
constructed by just "flipping" the axes of the CDF.  This is why the quantile
function is also called the inverse CDF.

::: example
**The whole quantile function**

We can use the same ideas as in the previous example to show that $F_{r}^{-1}(q)$ is
equal to $0$ for any $q$ between $0$ and $19/37$, and equal to $1$ for any $q$
between $19/37$ and $1$. But what is the value of $F^{-1}_{r}(19/37)$?  To
figure that out we will need to carefully apply the definition:
  \begin{align}
    F_{r}^{-1}(19/37) &= \min\{a \in S_x: F_{r}(a) \geq 19/37\} \\
      &= \min \{-1,1\} \\
      &= -1
  \end{align}
So the full quantile function can be written in case form as:
  \begin{align}
    F_{r}^{-1}(q) &= \begin{cases}
                    0 & 0 < q \leq 19/37 \\
                    1 & 19/37 < q < 1 \\
                  \end{cases}
  \end{align}
and we can plot it below:  
```{r RouletteQuantFunction, fig.cap = "*Quantile function for $r$*"}
RouletteQuant <- tibble(a = c(0.0001,19/37,0.9999),
                        qred = c(0,1,1))
ggplot(data = RouletteQuant, mapping = aes(x = a)) +
  geom_step(aes(y=qred),col = "red") +
  xlab("quantile") +
  ylab("value of r at this quantile") +
  geom_text(x = 0.8, 
            y = 0.9, 
            label = "q_r(a)", 
            col = "red") +
  labs(title = "Quantile function (inverse CDF)", 
       subtitle = "Roulette - indicator for red wins (r)", 
       caption = "", 
       tag = "")
```
Notice that this looks just like the CDF, but with the horizontal and vertical
axes flipped.
:::

### Median {#the-median}

The ***median*** of a random variable is its 0.5 quantile or 50th percentile.

::: example
**The median in roulette**

The median of $r$ is just its 0.5 quantile or 50th percentile:
  $$median(r) = F_{r}^{-1}(0.5) = 0$$
:::

Like the expected value, the median is often interpreted as a measure of central
tendency for the random variable.

## Functions of a random variable

Any function of a random variable is also a random variable.  So for example, if
$x$ is a random variable, so is $x^2$ or $\ln (x)$ or $\sqrt{x}$.

::: example
**The net profit from a bet on red ($y_{Red}$)**

The player's net profit from \$1 bet on red $(y_{Red})$ was earlier defined
directly from the underlying outcome $\omega$.
\begin{align}
  y_{Red} &= \begin{cases}  
      1 & \textrm{ if } \omega \in RedWins \\
      -1 & \textrm{ if } \omega \in RedWins^c 
      \end{cases} \\
\end{align}

Alternatively, we can define it in terms of the random variable $b$, or of the
random variable $r$:
\begin{align}
  y_{Red} &= 2I(b \in RedWins) - 1 \\
  y_{Red} &= 2r -1
\end{align}
where $I(\cdot)$ is the indicator function.
:::

When $y$ is a function of $x$, we can derive the probability distribution of
$y$ directly from the probability distribution of $x$, using the same method
we used in Section \@ref(probability-distributions). We can also use a similar
formula:
\begin{align}
  E( h(x) ) &= \sum_{s \in S_x} h(s) f_x(s)
\end{align}
to calculate the expected value of $y = h(x)$ from the PDF of $x$.

::: example
**The expected value of $y_{14}^2$**

Suppose we are interested in the expected value of $y_{14}^2$ - it's not obvious
why we would be, but we will be.

Applying the formula above we get:
\begin{align}
  E(y_{14}^2) &= \sum_{s \in S_{14}} s^2 f_{14}(s) \\
    &= \sum_{s \in \{-1, 35\}} s^2 f_{14}(s) \\
    &= (-1)^2 f_{14}(-1) + 35^2 f_{14}(35) \\
    &= 1 * 36/37 + 1225 * 1/37 \\
    &\approx 34.08
\end{align}
This result will turn out to be useful at some point.
:::

We say that $y$ is a ***linear function*** of $x$ when we can write it in the
form:
\begin{align}
  y &= a + bx
\end{align}
for some constants $a$ and $b$. When $y$ is a function of $x$ but cannot be
written in this form, we say that it is a ***nonlinear function*** of $x$.

::: example
**Linear and nonlinear functions in roulette**

The player's net profit from a bet on red is a linear function of the random
variable $r$:
\begin{align}
  y_{Red} = 2r -1
\end{align}
and it is a nonlinear function of the random variable $b$:
\begin{align}
  y_{Red} = 2I(b \in RedWins) - 1
\end{align}
:::

Linear functions are particularly convenient to work with.  In particular, 
we can prove that the expected value of any linear function of any random
variable $x$ is:
\begin{align}
  E(a + bx) &= a + b E(x)
\end{align}
That is, we do not need to know the entire distribution of $x$ to calculate 
$E(y)$; we only need to know the expected value $E(x)$.

::: example
**The expected value of $y_{Red}$**

Earlier, we showed that $y_{Red}$ can be defined as a linear function of $r$:
  $$y_{Red} = 2r -1$$
so its expected value can be derived:
  \begin{align}
    E(y_{Red}) &= E(2r - 1) \\
      &= 2 \underbrace{E(r)}_{18/37} - 1 \\
      &= -1/37 \\
      &\approx -0.027
  \end{align}
We can interpret this result as saying that the house advantage for bets on 
red is 2.7 cents per dollar bet. Note that this is the same house advantage
as we earlier derived for a bet on 14. 
:::

Unfortunately, this handy property applies only to linear functions. If
$h(\cdot)$ is a nonlinear function, than $E(h(x)) \neq h(E(x))$. Students
frequently make this mistake, so try to avoid it.

::: example
**The expected value of a nonlinear function in roulette**

We can define $y_{Red}$ as a nonlinear function of $b$:
  $$y_{Red} = 2 I(b \in RedWins) - 1$$
Can we take the expected value "inside" of this function?  That is, does
$E(y_{Red}) = 2 I(E(b) \in RedWins) - 1$?

The answer to this question is "No". We already showed that
$E(y_{Red}) \approx -0.027$. We also showed earlier that
$E(b) = 18$, so we can find:
  \begin{align}
    2 I(E(b) \in RedWins) - 1 &= 2 I(18 \in RedWins) - 1 \\
      &= 2*1 - 1 \\
      &= 1
  \end{align}
Since $-0.027 \neq 1$, it is clear that $E(y_{Red})$ is *not* equal to
$2 I(E(b) \in RedWins) - 1$.

More generally, there is *no* formula that would allow you to calculate the
expected value $E(y_{Red})$ using only the expected value $E(b)$.  You need the
whole distribution of $b$ to calculate $E(y_{Red})$.
:::


::: {.fyi data-latex=""}
**The expected value is a sum!**

The result that $E(a + bx) = a + b E(x)$ is one that we will use repeatedly.
While proving it is not something I would ask you to do, working through the
proof is useful in understanding why it works, and why the result only applies
to linear functions of $x$.

The key to all of this is that the expected value is a *sum*. Consider the
formula for the expected value above:
\begin{align}
  E( h(x) ) &= \sum_{s \in S_x} h(s) f_x(s)
\end{align}
When $y = h(x)$ is a linear function of $x$, this reduces to:
\begin{align}
  E(y) &= E( a + bx  ) \\
    &= \sum_{s \in S_x} (a + bs) f_x(s) 
\end{align}
At this point, remember three rules of arithmetic you learned in grade school.
The first rule is the distributive property of multiplication, which
says that $a*(b+c) = a*b + a*c$.  This allows us to rearrange:
\begin{align}
  E(y) &= \sum_{s \in S_x} (a + bs) f_x(s) \\
    &= \sum_{s \in S_x} a f_x(s) + bs f_x(s)
\end{align}
The other two rules are that you can rearrange a sum in any order you like (this
is called the commutative property of addition: $a + b = b + a$), and that
you can group sums in any way you like (the associative property of addition: 
$(a + b) + c = a + (b + c)$).  This allows us to rearrange the summations like
this:
\begin{align}
  E(y) &= \sum_{s \in S_x} a f_x(s) + bs f_x(s) \\
    &= \sum_{s \in S_x} a f_x(s) + \sum_{s \in S_x} bs f_x(s)
\end{align}
Then we apply the commutative property again:
\begin{align}
  E(y) &=\sum_{s \in S_x} a f_x(s) + \sum_{s \in S_x} bs f_x(s) \\
    &= a \underbrace{\sum_{s \in S_x} f_x(s)}_{= 1}  + b \underbrace{\sum_{s \in S_x} s f_x(s)}_{=E(x)} \\
    &= a + bE(x)
\end{align}
to get our result. 
:::

## Variance and standard deviation {#variance-and-standard-deviation}

In addition to measures of central tendency such as the expected value and
median, we are also interested in measures of "spread" or variability. We have
already seen one - the range - but there are others, including the variance and
the standard deviation.

### Variance {#variance}

The ***variance*** of a random variable $x$ is defined as:
$$\sigma_x^2 = var(x) = E((x-E(x))^2)$$
where $\sigma$ is the lower-case Greek letter *sigma*.


Variance can be thought of as a measure of how much $x$ tends to deviate from
its central tendency $E(x)$.

::: example
**Calculating variance from the PDF**

We can calculate the variance of $r$ directly from its PDF:
\begin{align}
  var(r) &= E( (r - E(r))^2 ) \\
    &= \sum_{s \in S_r} (s - E(r) )^2 f_r(s) \\
    &= \sum_{s \in \{0, 1\}} (s - E(r) )^2 f_r(s) \\
    &= (0 - E(r) )^2 f_r(0) + (1 - E(r) )^2 f_r(1) \\
    &\approx (0 - 0.486 )^2 * 0.514 + (1 - 0.486 )^2 * 0.486 \\
    &\approx 0.25
\end{align}
Note that we used the formula for the expected value of a function of 
a random variable, then substituted, expanded the sum, and substituted
again to get the result.
:::

The key to understanding the variance is that it is the expected value of
a square $(x-E(x))^2$, and the expected value is just a (weighted) sum.
This has several implications:

1. The variance is always *positive* (or more precisely, non-negative):
     $$var(x) \geq 0$$
   The intuition for this result is straightforward: the square of any number
   is always positive, and the expected value is just a sum.  The variance is
   therefore a sum of several positive numbers, so it is also a positive number.
2. The variance can also be written in the form:
     $$var(x) = E(x^2) - E(x)^2$$
   The derivation of this is as follows:
   \begin{align}
     var(x) &= E((x-E(x))^2) \\
       &= E( ( x-E(x) ) * (x - E(x) )) \\
       &= E( x^2 - 2xE(x) + E(x)^2)  \\
       &= E(x^2) - 2E(x)E(x) + E(x)^2 \\
       &= E(x^2) - E(x)^2
   \end{align}
   This formula is often an easier way of calculating the variance.

::: example
**Calculating variance using the alternate formula**

We can use the alternative formula to calculate $var(y_{14})$. 

1. Calculate $E(y_{14})$ from the PDF of $y_{14}$. We found this earlier:
   \begin{align}
      E(y_{14}) &= -0.027
    \end{align}
2. Calculate $E(y_{14}^2)$ from the PDF of $y_{14}$. We also found this
   earlier:
   \begin{align}
     E(y_{14}^2) &\approx 34.08
   \end{align}
4. Put these two results into the formula:
   \begin{align}
     var(y_{14}) &= E(y_{14}^2) - E(y_{14})^2 \\
       &\approx 34.08 + (-0.027)^2 \\
       &\approx 34.1
    \end{align}

This is the same result you would get if you calculated $var(y_{14})$ directly
from the PDF, but you may find it easier to calculate.
:::

3. We can also find the variance of any linear function of a random variable. For
   any constants $a$ and $b$:
     $$var(a + bx) = b^2 var(x)$$
   This can be derived as follows:
   \begin{align}
     var(a+bx) &= E( ( (a+bx) - E(a+bx))^2) \\
       &= E( ( a+bx - a-bE(x))^2) \\
       &= E( (b(x - E(x)))^2) \\
       &= E( b^2(x - E(x))^2) \\
       &= b^2 E( (x - E(x))^2) \\
       &= b^2 var(x)
    \end{align}

::: example
**Calculating the variance of a linear function**

We earlier found that $var(r) \approx 0.25$, so we can find the variance of 
$y_{Red}$ using our formula for the variance of a linear function:
  \begin{align}
    var(y_{Red}) &= var( 2r - 1) \\
      &= 2^2 var(r) \\
      &\approx 4*0.25 \\
      &\approx 1.0
  \end{align}
:::

The variance has a natural interpretation as a measure of how much the random
variable tends to vary.  Low variance means you get similar results each time,
high variance means that results vary a lot.

::: example
**Interpreting the variance of $y_{Red}$ and $y_{14}$**

We earlier found that a bet on red and a bet on 14 have the same expected value:
\begin{align}
  E(y_{red}) &\approx -0.027 \\
  E(y_{14}) &\approx -0.027
\end{align}
However, a bet on red has a much lower variance:
\begin{align}
  var(y_{red}) &\approx 1.0 \\
  var(y_{14}) &\approx 34.1
\end{align}
What this means in the context of betting is that all players will tend to lose
money on average with both bets (low and negative expected value) but most
players who bet on red will tend to have similar results (everyone loses their
money gradually) while players who bet on 14 will tend to have more variable
results (a few players will have a big win or two, while most players lose their
money quickly).
:::

### Standard deviation {#standard-deviation}

The ***standard deviation*** of a random variable is defined as the (positive)
square root of its variance:
$$\sigma_x = sd(x) = \sqrt{var(x)}$$
The standard deviation is just another way of describing the variability of $x$.  

In some sense, the variance and standard deviation are interchangeable since
they are so closely related.  The standard deviation has the advantage that it
is expressed in the same units as the underlying random variable, while the
variance is expressed in the square of those units.  This makes the standard
deviation somewhat easier to interpret.

::: example
**Standard deviation in roulette**

The standard deviation of $r$ is:
  $$sd(r) = \sqrt{var(r)} \approx \sqrt{0.25} \approx 0.5$$

The standard deviation of $y_{Red}$ is:
  $$sd(y_{Red}) = \sqrt{var(y_{Red})} \approx \sqrt{1.0} \approx 1.0$$
  
The standard deviation of $y_{14}$ is 
  $$sd(y_{14}) = \sqrt{var(y_{14})} \approx \sqrt{34.1} \approx 5.8$$
:::

The standard deviation has analogous properties to the variance:

1. It is always non-negative:
   $$sd(x) \geq 0$$
2. For any constants $a$ and $b$:
   $$sd(a + bx) = b \, sd(x)$$

These properties follow directly from the corresponding properties of the
variance.

### Standardization {#standardization}

In some cases, it is useful to ***standardize*** a random variable. This means
constructing a new random variable of the form:
\begin{align}
  z &= \frac{x - E(x)}{sd(x)}
\end{align}
where $x$ is the original random variable and $z$ is the standardized version
of $x$. That is, we are subtracting the expected value, and dividing by the
standard deviation.

Standardization is an example of a change in units, just like converting 
miles to kilometers, or grams to kilograms.  Like most changes in units, it
is a *linear* transformation.  That is, we can rewrite $z$ as a linear function
of $x$:
\begin{align}
  z &= \frac{-E(x)}{sd(x)} + \frac{1}{sd(x)} * x
\end{align}
We can also use our rules for linear functions of a random variable to get:
\begin{align}
  E(z) &= \frac{-E(x)}{sd(x)} + \frac{1}{sd(x)} * E(x) \\
    &= 0 \\
  var(z) &= \left(\frac{1}{sd(x)}\right)^2 * var(x) \\
    &= 1
\end{align}
That is, standardization rescales the variable to have a mean of zero and 
a variance or standard deviation of one.

Standardization is commonly used in fields like psychology or educational
testing when a variable has no natural unit of measurement.

::: example
**A standardized test score**

Suppose that the midterm exam in this course is graded on a scale from 0 to 100,
with a mean score of $E(x) = 70$ and a standard deviation of $sd(x) = 10$. For
any individual student's score $x$, the standardized score is:
  $$ z = \frac{x-70}{10} $$
or (equivalently):
  $$z = 0.1  x - 7 $$
Applying our results on linear functions of a random variable:
  \begin{align}
    E(z) &= 0.1 E(x) - 7 \\
      &= 0.1 \times 70 - 7 \\
      &= 0 \\
    sd(z) &= 0.1  sd(x) \\
      &= 0.1 \times 10 \\
      &= 1
  \end{align}
Students with a positive standardized test score $(z > 0)$ did better than
average, while students with a negative standardized score did worse than
average.
:::

## Standard discrete distributions {#standard-distributions}

In principle, there are an infinite number of possible probability
distributions.  However, some probability distributions appear so often in
applications that we have given them names. This provides a quick way to
describe a particular distribution without writing out its full PDF, using the
notation:
  $$RandomVariable \sim DistributionName(Parameters)$$
where $RandomVariable$ is the name of the random variable whose distribution is
being described, the $\sim$ character can be read as "has the following
probability distribution", $DistributionName$ is the name of the probability
distribution, and $Parameters$ is a list of arguments called ***parameters***
that provide additional information about the probability distribution.

Using a standard distribution also allows us to establish the properties of a
commonly-used distribution once, and use those results every time we use that
distribution.  In this section we will describe three standard distributions - 
the Bernoulli, the binomial, and the discrete uniform - and their properties.

### Bernoulli {#bernoulli}

The ***Bernoulli*** probability distribution is usually written:
$$x \sim Bernoulli(p)$$
It has discrete support $S_x = \{0,1\}$ and PDF:
\begin{align}
  f_x(a) &= \begin{cases}
    (1-p) & \textrm{if $a = 0$} \\
    p & \textrm{if $a = 1$} \\
    0 & \textrm{otherwise}\\
    \end{cases}
\end{align}
Note that the "Bernoulli distribution" isn't really a (single) probability
distribution.  Instead it is what we call a ***parametric family*** of
distributions.  That is, the $Bernoulli(p)$ is a different distribution with a
different PDF for each value of the ***parameter*** $p$.

We typically use Bernoulli random variables to model the probability of some
random event $A$. If we define $x$ as the indicator variable $x=I(A)$, then 
$x \sim Bernoulli(p)$ where $p=\Pr(A)$.

::: example
**The Bernoulli distribution in roulette**

The variable $r = I(\omega \in RedWins)$ has the $Bernoulli(18/37)$ distribution.
:::

The mean of a $Bernoulli(p)$ random variable is:
\begin{align}
  E(x) &= (1-p)*0 + p*1 \\
       &= p
\end{align}
and its variance is:
\begin{align}
var(x) &= E(x^2) - E(x)^2 \\
  &= (0^2*(1-p) + 1^2 p) - (p)^2 \\
  &= p - p^2
\end{align}
If you recognize that a particular random variable has a Bernoulli distribution,
you can use these two results to save yourself the trouble of doing the
calculations by hand.

### Binomial {#binomial}

The ***binomial*** probability distribution is usually written:
$$x \sim Binomial(n,p)$$
It has discrete support $S_x = \{0,1,2,\ldots,n\}$ and its PDF is:
$$f_x(a) = 
  \begin{cases} 
    \frac{n!}{a!(n-a)!} p^a(1-p)^{n-a} & \textrm{if $a \in S_x$} \\ 
    0 & \textrm{otherwise} \\ 
  \end{cases}$$
You do not need to memorize or even understand this formula.  The Excel function 
`BINOMDIST()` can be used to calculate the PDF or CDF of the binomial
distribution, and the function `BINOM.INV()` can be used to calculate its
quantiles.

The binomial distribution is typically used to model frequencies or counts. We
can show that it is the distribution of how many times a probability-$p$ event
happens in $n$ independent attempts.

::: example
**The binomial distribution in roulette**

Suppose we play 50 (independent) games of roulette, and bet on red in every
game. Since the outcome of a single bet on red is $r \sim Bernoulli(18/37)$,
the number of times we win is:
  $$WIN50 \sim Binomial(50,18/37)$$
We can use the Excel formula `=BINOM.DIST(25,50,18/37,FALSE)` to calculate the
probability of winning exactly 25 times:
  $$\Pr(WIN50 = 25) \approx 0.11$$
we can use the Excel formula `= BINOM.DIST(25,50,18/37,TRUE)` to
calculate the probability of winning 25 times or less:
  $$\Pr(WIN50 \leq 25) \approx 0.63$$  
and we can use the Excel formula `= 1 - BINOM.DIST(25,50,18/37,TRUE)` to
calculate the probability of winning more than 25 times:
  $$\Pr(WIN50 > 25) = 1 - \Pr(WIN50 \leq 25) \approx 0.37$$
So we have a 37\% chance of making money (winning more often than losing),
an 11\% chance of breaking even, and a 52\% chance of losing money. Notice that
the chance of losing money over the course of 50 games (52%) is greater than
the chance of losing money in a single game (51.4%).  If you play enough games,
the chance of losing money becomes very close to 100%.
:::

The mean and variance of a binomial random variable are:
  $$E(x) = np$$
  $$var(x) = np(1-p)$$
Again, if you recognize a random variable as having a binomial distribution
you can use these results to save yourself the trouble of doing the
calculations by hand.

::: example
**The binomial distribution in roulette, part 2**

The number of wins in 50 bets on red has expected value:
  $$E(WIN50) = np = 50 * 18/37 \approx 24.3$$
variance:
  $$var(WIN50) = np(1-p) = 50 * 18/37 * 19/37 \approx 12.5$$
and standard deviation:
  $$sd(WIN50) = \sqrt{var(WIN50)} \approx \sqrt{12.5} \approx 3.5$$
:::

::: {.fyi data-latex=""}
The formula for the binomial PDF looks strange, but it can actually be derived
from a fairly simple and common situation.  Let our outcome
$\omega = (b_1,b_2,\ldots,b_n)$ be a sequence of $n$ independent random
variables from the $Bernoulli(p)$ distribution and let:
  $$x = \sum_{i=1}^n b_i$$
count up the number of times that $b_i$ is equal to one (i.e., the event modeled
by $b_i$ happened).  Then it is possible to derive the PDF for $y$, and that is
the PDF we call $Binomial(n,p)$.  The derivation is not easy, but the intuition
is simple: 

- The number of distinct outcomes in the event $x=a$ is $\frac{n!}{a!(n-a)!}$.
- The probability of each individual outcome in the event $x=a$ is
  $p^a(1-p)^{n-a}$.
- We can calculate the probability of the event $x=a$ by adding up these
  probabilities.

Therefore the probability of the event $x=a$ is
$\frac{n!}{a!(n-a)!}p^a(1-p)^{n-a}$.
:::  

### Discrete uniform {#discrete-uniform}

The ***discrete uniform*** distribution:
  $$x \sim DiscreteUniform(S_x)$$
is a distribution that puts equal probability on every value in a discrete set
$S_x$. Its support is $S_x$ and its PDF is:
$$f_x(a) = \begin{cases}
          1/|S_x| & a \in S_x \\
          0 & a \notin S_x \\
          \end{cases}$$
Discrete uniform distributions appear in gambling and similar applications.  

::: example
**The discrete uniform distribution in roulette**

In our roulette example, the random variable $b$ has the
$DiscreteUniform(\{0,1,\ldots,36\})$ distribution.
:::

## Chapter review {-#review-random-variables}

Random variables are simply numerical random outcomes. Economics is a primarily
quantitative field, in the sense that most outcomes we study are numbers.
This quantification helps us be more precise in our analysis and predictions.

In this chapter we have learned various ways of describing the probability 
distribution of a simple random variable - a single random variable that 
takes on values in a finite set.  We have also learned some standard probability
distributions for simple random variables.

Later in the course we will deal with
[more complex random variables](#more-on-random-variables)
including random variables that take on values in a continuous set, as well as
pairs or groups of related random variables.  We will then apply the concept of
a random variable to [statistics](#statistics) calculated from data.

## Practice problems {-#problems-random-variables}

Answers can be found in the [appendix](#answers-random-variables).

The questions below continue our [craps example](#problems-probability). To 
review that example, we have an outcome $(r,w)$ where $r$ and $w$ are the 
numbers rolled on a pair of fair six-sided dice.

Let the random variable $t$ be the total showing on the pair of dice, and let
the random variable $y = I(t=11)$ be an indicator for whether a bet on "Yo" 
wins.

**GOAL #1: Define a random variable in terms of a random outcome**

1. Define $t$ in terms of the underlying outcome $(r,w)$.

2. Define $y$ in terms of the underlying outcome $(r,w)$.

**GOAL #2: Determine the support and range of a random variable**

3. Find the support of the following random variables:
   a. Find the support $S_r$ of the random variable $r$.
   b. Find the support $S_t$ of the random variable $t$.
   c. Find the support $S_y$ of the random variable $y$.

4. Find the range of each of the following random variables:
   a. Find the range of $r$.
   b. Find the range of $t$.
   c. Find the range of $y$.

**GOAL #3: Calculate and interpret the PDF of a discrete random variable**

5. Find the following PDFs:
   a. Find the PDF $f_r$ for the random variable $r$.
   b. Find the PDF $f_t$ for the random variable $t$.
   c. Find the PDF $f_y$ for the random variable $y$.

**GOAL #4: Calculate and interpret the CDF of a discrete random variable**

6. Using the PDFs you found earlier, find the following CDFs:
   a. Find the CDF $F_r$ for the random variable $r$.
   b. Find the CDF $F_y$ for the random variable $y$.

**GOAL #5: Calculate interval probabilities from the CDF**

7. Suppose that the discrete random variable $x$ has CDF $F_x$ where
   $F_x(0) = 0.3$, $F_x(5) = 0.8$, $f_x(0) = 0.1$, and $f_x(5) = 0.1$. Find the
   following interval probabilities:
   a. Find $\Pr(x \leq 5)$.
   b. Find $\Pr(x < 5)$.
   c. Find $\Pr(x > 5)$.
   d. Find $\Pr(x \geq 5)$.
   e. Find $\Pr(0 < x \leq 5)$.
   f. Find $\Pr(0 \leq x \leq 5)$.
   g. Find $\Pr(0 < x < 5)$.
   h. Find $\Pr(0 \leq x < 5)$.

**GOAL #6: Calculate the expected value of a discrete random variable from its PDF**

8. Using the PDFs you found earlier, find the following expected values:
   a. Find the expected value $E(r)$.
   b. Find the expected value $E(r^2)$.

**GOAL #7: Calculate a quantile from the CDF**

9. Using the CDFs you found earlier, find the following quantiles:
   a. Find the median $Med(r)$.
   b. Find the 0.25 quantile $F_r^{-1}(0.25)$.
   c. Find the 75th percentile of $r$.

**GOAL #8: Calculate the variance of a discrete random variable from its PDF**

10. Let $d = (y - E(y))^2$.
    a. Find the PDF $f_d$ of $d$.
    b. Use this PDF to find $E(d)$
    c. Use these results to find the variance $var(y)$.

**GOAL #9: Calculate the variance from expected values**

11. In question (8) above, you calculated $E(r)$ and $E(r^2)$ from the PDF. Use
    these results to find $var(r)$.

**GOAL #10: Calculate the standard deviation from the variance**

12. Find the following standard deviations:
    a. Find $sd(y)$. You can use your result from question (10) above.
    b. Find $sd(r)$. You can use your result from question (11) above.

**GOAL #11: Calculate the expected value for a linear function of a random variable**  
**GOAL #12: Calculate variance and standard deviation for a linear function of a random variable**

13. The "Yo" bet pays out at 15:1, meaning you win \$15 for each dollar bet. Suppose
    you bet \$10 on Yo.  Your net winnings in that case will be $W = 160*y - 10$.
    a. Using earlier results, find $E(W)$.
    b. Using earlier results, find $var(W)$.
    c. The event $W > 0$ (your net winnings are positive) is identical to the
       event $y = 1$.  Using earlier results, find $\Pr(W > 0)$.

14. Suppose you bet \$1 on Yo in ten independent rolls. Your net winnings in 
    that case will be $W_{10} = 16*Y_{10} - 10$.
    a. Using earlier results, find $E(W_{10})$.
    b. Using earlier results, find $var(W_{10})$.
    c. The event $W_{10} > 0$ (your net winnings are positive) is identical to
       the event $Y_{10} > 10/16$.  Using earlier results, find
       $\Pr(W_{10} > 0)$.

**GOAL #13: Standardize a random variable**

15. Let $z$ be the standardized form of the random variable $y$.
    a. What is the formula defining $z$? Use actual numbers for $E(y)$ and
       $sd(y)$.
    b. Find the support of $z$.
    c. Find the PDF of $z$.

**GOAL #14: Use standard standard discrete probability distributions**

16. The random variable $y$ can be described using a standard distribution.
    a. What standard distribution describes $y$?
    b. Use standard results for this distribution to find $E(y)$
    c. Use standard results for this distribution to find $var(y)$

17. Let $Y_{10}$ be the number of times in 10 dice rolls that a bet on "Yo"
    wins.
    a. What standard distribution describes $Y_{10}$?
    b. Use existing results for this distribution to find $E(Y_{10})$.
    c. Use existing results for this distribution to find $var(Y_{10})$.
    d. Use Excel to calculate $\Pr(Y_{10} = 0)$.
    e. Use Excel to calculate $\Pr(Y_{10} \leq 10/16)$.
    f. Use Excel to calculate $\Pr(Y_{10} > 10/16)$.
