# Advanced data analysis

```{r setup11, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

In this chapter we will learn how to:

- Implement and interpret a simple hypothesis test.
- Calculate and interpret common multivariate statistics such as covariance 
  and correlation.
- Construct and interpret a pivot table.
- Construct and intepret a scatter plot.


## Multivariate statistics

### Scatter plots 

Figure ??? below is an example of a ***scatter plot***. We use scatter plots
to visualize the relationship between two variables in a data set. Each
observation is represented by a single point.

```{r}
dat <- tibble(xval=rnorm(100),yval=1+0.5*xval+0.2*rnorm(100))
ggplot(data=dat, mapping=aes(x=xval,y=yval)) +
    geom_point() + ggtitle("An example of a scatter plot")
```

#### Creating a scatter plot in Excel

To create a scatter plot in Excel:

1. Select the data you want to use in your scatter plot.
    - In this example, it will be [add instructions here]
2. Select `Insert > Recommended Charts` from the menu.
3. The `Insert Chart` dialog box will show a few suggested charts, but
  let's ignore those.  Instead click on the tab at the top marked
  `All charts`.
4. The dialog box will now show a large number of chart types. 
    - While we are here, take a look at these options for future reference.
    - The one we want is called "X Y (Scatter)", so click on that button.
5. You will see two options. Click the one on the right, and then click "OK".

Now we have a scatter plot.  

#### Scatter plot design



#### Modifying your scatter plot

Although we have created our scatter plot, there are many ways it
isn't exactly what we might like.  

Let's start by changing the title of the graph from its default 
of "Chart Title":

1. Double-click on "Chart Title" 
2. Type in the revised title [insert title here].
3. Click somewhere else.
  
In general if you want to change someting already in the graph, click 
or double-click on it.

Now let's add some axis labels to tell our viewer what these numbers refer
to.  

1. Click on the chart. You should now be seeing the `Design` ribbon.
2. Click on the `Add Chart Element` button on the Design ribbon. Select
  `Axis Titles` and then `Primary Horizontal`.
    - You should now see the title "Axis title" on the horizontal 
      axis. 
3. Double click on "Axis title" to change the axis title to "Income"
4. Repeat the steps 2-3 above to add the vertical axis title "Spending."

We can also use the `Add Chart Element` button to remove elements, or we
can just click on them and press the `Delete` key.

As with most other objects, you can resize, move, copy-and-paste, and 
delete Excel graphs.

### Sample covariance and correlation

To calculate the covariance of two variables, use the `COVARIANCE.S()` 
function.

To calculate the correlation of two variables, use the `CORREL()` 
function.

### Conditional averages

You can use the `AVERAGEIFS()` function to calculate 
the conditional average of one variable given a condition on another 
variable (or even the same variable).

These functions both require three pieces of information:

- The variable to average
- The variable used to calculate the condition.
- The condition for inclusion in the average

For example, suppose we want to calculate the average income of men
in our data:

- The variable to average is Income (cells C2:C5)
- The variable used to calculate the condition is Gender (cells B2:B5).
- The condition is "Male"

so go to cell ?? and enter `=AVERAGEIFS(C2:C5,B2:B5,"Male")`  The cell
will now display the average income of males in our data.

The `AVERAGEIF()` function can also be used for this purpose, but
its format is slightly different and it is less powerful than 
`AVERAGEIFS()`.

### Pivot tables

Pivot tables are quite powerful but somewhat tricky to use. 

#### A simple pivot table

Let's start by creating a simple pivot table

1. Click anywhere in your data table.
2. Select `Insert` and then `PivotTable` 
3. Excel will display a dialog box allowing you to specify the range of
  your data along with a few other options.  THe default here is fine,
  so click on `OK`.
  
You will now see something like this:

> INSERT SCREENSHOT HERE.

Let's start with a simple one-variable table. Check the box next to "Gender,"
and you will see something like this:

> INSERT SCREENSHOT HERE.

We now can report various statistics broken down by gender.  Let's start with 
a simple count.  Drag "Gender" into the box marked "$\Sigma$ values". You
will see something like this:

> INSERT SCREENSHOT HERE.

As we can see, it shows the number of observations for each value of Gender.
Suppose we want to change it from a count to a percentage.  Then right-click
on "Count of Gender", then on "Show Values As" and then on "% of column total."
You will now see a screen like this:

> INSERT SCREENSHOT HERE.

Now let's see what income by gender looks like.  Drag "Income" into the 
box marked "$\Sigma$ values".  You will now see a screen that looks like 
this:

> INSERT SCREENSHOT HERE.

Unfortunately, we wanted to see the average income by gender, but instead
we see the sum of income for each gender.  We can change that by right-clicking
on "Sum of Income", then on "Summarize Values By" and then "Average".

We can keep adding variables, and ways to summarize them. But there
are other things we can do.

#### Improving our pivot table

- We can sort rows by any column in the table.
- We can filter rows out of the table:
  - We can do this on our row label by clicking on the first column header
    and selecting the filter we want.
  - We can also filter on any other variable by dragging that variable name
    into the "Filters" box.

## Connecting theory and data

### Probability distributions

Excel includes many functions for calculating the PDF, CDF and inverse CDF
of various standard probability distributions. 

- The `NORM.S.DIST()` function calculates the PDF or CDF of the standard
  normal or $N(0,1)$ distribution.  It takes two arguments:
  - `Z` is the value at which to report the PDF or CDF.
  - `Cumulative` is a logical value that tells Excel whether to 
    report the CDF (if `TRUE`) or the PDF (if `FALSE`).

  For example, `NORM.S.DIST(0,TRUE)` returns $\Phi(0)$ or 0.5,
  while `NORM.S.DIST(0,FALSE)` returns $\phi(0)$ or about 0.399.
- The `NORM.S.INV()` function calculates the inverse CDF of the standard
  normal distribution.  It takes one argument:
  - `Probability` is the quantile (between 0 and 1) or probability 
    at which to calculate the inverse CDF or quantile function.  
  
  For example,`NORM.S.INV(0.95)` reports the 95th percentile of 
  the $N(0,1)$ distribution, or about 1.645.
- `NORM.DIST()` and `NORM.INV()` are similar functions for the general
  normal $N(\mu,\sigma)$ distribution.
- `T.DIST()` and `T.INV()` are similar functions for the $T_{n-1}$ 
  distribution.

### Simulation 

  - RAND (generates U(0,1) random number)


### Statistical inference

#### Finite-sample inference on the mean

Let $x_i$ be the value of the `Income` variable in our data.

Let $\mu_x = E(x_i)$. Suppose we want to test the null hypothesis:
  $$H_0: \mu_x = 30000$$
against the alternative:
  $$H_1: \mu_x \neq 30000$$
at a significance level of 5\%. 

1. Open a new worksheet and name it "Finite sample inference"
2. Write the following headings in the first row, starting with cell A1:

| Ho | Size | n | xbar | s_x | T | cL | cH | A/R | p-value |CI_low | CI_high |
|----|------|---|------|-----|---|----|----|-----|---------|-------|---------|

2. Set the characteristics of the test we want to perform:
    - Fill in cell A2 with 30000, our value of $\mu_x$ under the null.
    - Fill in cell B2 with 0.05, our desired significance level under the null.
3. Calculate our test statistic:
    - Use Excel's `COUNT()` function to fill in cell C2 with the number of
      (numeric) observations on `Income` in our data set.
    - Use Excel's `AVERAGE()` function to fill in cell D2 with the average 
      value of `Income` in our data set.
    - Use Excel's `STDEV.S()` function to fill in cell E2 with the standard 
      deviation of `Income` in our data set.
    - Calculate the test statistic in cell F2, using the statistics you
      have already calculated.
4. Calculate the critical values:
    - Use Excel's `T.INV()` function to calculate the critical values 
      for your test from the $t_{n-1}$ distribution in cells G2 and H2.
    - Use Excel's `IFS()` function to display "Accept" or "Reject"
      based on your test statistic and critical values in cell I2.
5. Calculate p-values:
    - I have not taught you the formula for a p-value, so I will just
      give it to you.
    - Enter `=T.DIST.2T(F2,C2-1)` in cell J2 to get the p-value of this test.
6. Calculate a confidence interval:
    - Use the results you have already calculated to generate a 95\% confidence
      interval in cells K2 and L2.

If you want to run a different test, you can change the null hypothesis 
in A2 or the significance level in B2, and the test will automatically update.

You can also copy everything in row 2 to another row (you will need to make
some cell references absolute rather than relative) and run additional tests.

#### Asymptotic inference on the mean

Next we will try the same result, but using asymptotic approximations.

1. Copy the "Finite sample inference" worksheet and name the
  new worksheet "Asymptotic inference".
  
We are already almost done! Remember that the only difference between
finite sample and asymptotic inference is the assumed distribution
of the test statistic under the null:

- In finite sample inference, the test statistic is (exactly) $T_{n-1}$
  under the null.
- In asymptotic inference, the test statistic is (approximately) $N(0,1)$
  under the null.

So all we need to do is:

1. Change the critical value calculation in cells G2 and H2 to use the correct
  distribution.  
  - The Excel function to calculate the inverse CDF of the $N(0,1)$ distribution 
    is `NORM.S.INV()`
2. Change the p-value calculation in cell J2 to use the correct distribution.
  - The Excel function to calculate the CDF of the normal distribution is
    `NORM.S.DIST()`
  - The specific formula you will want is `=2*(1-NORM.S.DIST(F2,TRUE))`

That's it, we have asymptotic inference.


