---
title: "Example R Markdown document for Using R chapter"
author: "Brian Krauth"
date: "9/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This R Markdown document contains all of the code in the chapter on 
Using R.  You can run it by opening it in RStudio and pressing the "Knit"
button.

# Read in and clean data

Load the tidyverse and read in the data
```{r}
library("tidyverse")
EmpData <- read_csv("EmploymentData.csv")
#EmpData <- read_csv("https://bookdown.org/bkrauth/BOOK/sampledata/EmploymentData.csv")
```

Then clean it

- Make permanent changes to EmpData
- Change MonthYr to date format
- Add UnempPct and LFPPct

```{r}
EmpData <- EmpData %>%
  mutate(MonthYr = as.Date(MonthYr, "%m/%d/%Y")) %>% 
  mutate(UnempPct = 100*UnempRate) %>% 
  mutate(LFPPct = 100*LFPRate)
```


# View data

Show selected variables for observations with unemployment rate 
over 12.5%, sorted by unemployment rate:
```{r}
EmpData %>% 
  filter(UnempPct > 12.5) %>%
  select(MonthYr, UnempPct, LFPPct, PrimeMinister) %>%
  arrange(UnempPct)
```

# Analyze data

Summary statistics for EmpData
```{r}
summary(EmpData)
```

Mean of UnempPct
```{r}
mean(EmpData$UnempPct)
```

Variance of UnempPct
```{r}
var(EmpData$UnempPct)
```

SD of UnempPct
```{r}
sd(EmpData$UnempPct)
```

Median of UnempPct
```{r}
median(EmpData$UnempPct)
```

Mean of AnnPopGrowth
```{r}
mean(EmpData$AnnPopGrowth)
```

Mean of AnnPopGrowth (dropping NA values)
```{r}
mean(EmpData$AnnPopGrowth, na.rm = TRUE)
```

Mean of each column
```{r}
EmpData %>%
  select(where(is.numeric)) %>%
  lapply(mean, na.rm = TRUE)
```

Standard deviation of each column
```{r}
EmpData %>%
  select(where(is.numeric)) %>%
  lapply(sd, na.rm = TRUE) 
```

Simple Frequency table of PrimeMinister:
```{r}
EmpData %>% 
  count(PrimeMinister)
```

Binned frequency table of UnempPct:
```{r}
EmpData %>% 
  count(cut_interval(UnempPct, 6))
```
# Create graphs

Histogram
```{r}
ggplot(data = EmpData,
       aes(x = UnempPct)) + 
  geom_histogram(binwidth = 0.5,
                 fill = "blue") +
  geom_density() +
  labs(title = "Unemployment rate",
      subtitle = paste("January 1976 - January 2021 (",
                     nrow(EmpData),
                     " months)",
                     sep = "",
                     collapse = ""),
      caption = "Source: Statistics Canada, Labour Force Survey",
      tag = "Canada") +
  xlab("Unemployment rate, %") +
  ylab("Count")
```

Time series (line) graph
```{r}
ggplot(data = EmpData,
       aes(x = MonthYr)) + 
  geom_line(aes(y = UnempPct),
            col = "blue") +
  geom_text(x = as.Date("1/1/2000", "%m/%d/%Y"),
            y = 15,
            label = "Unemployment",col="blue") +
  geom_line(aes(y = LFPPct),
            col = "red") +
  geom_text(x = as.Date("1/1/2000", "%m/%d/%Y"),
            y = 60,
            label = "LFP",
            col = "red") +
  labs(title = "Unemployment and LFP rates",
      subtitle = paste("January 1976 - January 2021 (",
                     nrow(EmpData),
                     " months)",
                     sep = "",
                     collapse = ""),
      caption = "Source: Statistics Canada, Labour Force Survey",
      tag = "Canada") +
  xlab("") +
  ylab("Percent")  
```
